---
title: "Education_analysis"
author: "Emma Spors & Nichole Peters"
date: "11/21/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = TRUE, message=FALSE, warning=FALSE}
## Load the libraries
library(tidytuesdayR)
library(ggplot2)
library(plotly)
library(dplyr)
library(tidyr)
library(corrplot)
library(corrgram)
library(usmap)
library(moments)
library(animation)
library(tidyverse)
library(ggpubr)
```

```{r, echo = TRUE, message=FALSE, warning=FALSE}
## Pull the datasets
tuition_cost <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/tuition_cost.csv')

tuition_income <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/tuition_income.csv') 

salary_potential <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/salary_potential.csv')

historical_tuition <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/historical_tuition.csv')

diversity_school <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/diversity_school.csv')
```

```{r}
#Modify year 
historical_tuition$yearmod <- sapply(strsplit(historical_tuition$year, split='-', fixed=TRUE), function(x) (x[1]))
historical_tuition$yearmod <- as.integer(historical_tuition$yearmod)

#Create a subset for All Institutions
historical_tuitionALL <- subset(historical_tuition, historical_tuition$type == "All Institutions")
historical_tuitionpublic <- subset(historical_tuition, historical_tuition$type == "Public")
historical_tuitionprivate <- subset(historical_tuition, historical_tuition$type == "Private")

#Create subsets for all, 2, and 4 year institutions
histall <- subset(historical_tuitionALL, historical_tuitionALL$tuition_type == "All Constant")
hist2year <- subset(historical_tuitionALL, historical_tuitionALL$tuition_type == "2 Year Constant")
hist4year <- subset(historical_tuitionALL, historical_tuitionALL$tuition_type == "4 Year Constant")
```

```{r}
## Plot tutition cost over time for all institutions
p <- ggplot(histall, aes(yearmod, tuition_cost)) +
  labs(x = "Year", y="Tuition Cost", title = "All Institution Cost over Time") +
  theme(axis.text.x = element_text(angle = 90))+
  geom_col()
ggplotly(p)
```

```{r}
## Plot tutition cost over time for 2 year institutions
p <- ggplot(hist2year, aes(yearmod, tuition_cost)) +
  labs(x = "Year", y="Tuition Cost", title = "2 Year Institution Cost over Time") +
  theme(axis.text.x = element_text(angle = 90))+
  geom_col()
ggplotly(p)
```

```{r}
## Plot tutition cost over time for 4 year institutions
p <- ggplot(hist4year, aes(yearmod, tuition_cost)) +
  labs(x = "Year", y="Tuition Cost", title = "4 Year Institution Cost over Time") +
  theme(axis.text.x = element_text(angle = 90))+
  geom_col()
ggplotly(p)
```

```{r, echo = TRUE, message=FALSE, warning=FALSE}}
  #create line graph to compare the historical trend of tuition costs across different universities 
hist_tuition <- historical_tuition %>%
  filter(tuition_type == "All Constant" | tuition_type == "4 Year Constant" | tuition_type == "2 Year Constant") %>%
  filter(type == "All Institutions")
ts_plot <- ggplot(data = hist_tuition, aes(x = year, group = tuition_type)) + 
  geom_line(aes(y = tuition_cost, color = tuition_type), size = 1) + 
  ggtitle("Historical Costs of College \n Adjusted for Inflation") + 
  theme(axis.text.x = element_text(angle = 90)) + 
  xlab("Year") + 
  ylab("College Tuition Cost") + 
  labs(color = "Type of Degree")
  
ggplotly(ts_plot)
```


Tuition costs have continued to increase over the years, especially in for 4 year institutions, even when controlling for inflation. 
From 1985 to 2016 the cost of 2 year institutions has gone up \$3,090 whereas the cost of 4 year institutions has risen $14,319. There's a significant difference in the increase of price between the two.
```{r}
## Get a summary of the data and check variable types
summary(histall)
```

```{r}
## Create a linear model for cost of tuition
modelall <- lm(tuition_cost ~ yearmod, data = histall)
modelall
model4year <- lm(tuition_cost ~ yearmod, data = hist4year)
model4year
model2year <- lm(tuition_cost ~ yearmod, data = hist2year)
model2year
```

```{r}
## Summary of the models to check for accuracy
summary(modelall)
summary(model4year)
summary(model2year)
```

```{r, echo = TRUE, message=FALSE, warning=FALSE}
## Plot the linear models alongside historical costs
hist_tuition <- historical_tuition %>%
  filter(tuition_type == "All Constant" | tuition_type == "4 Year Constant" | tuition_type == "2 Year Constant") %>%
  filter(type == "All Institutions")
ts_plot <- ggplot(data = hist_tuition, aes(x = yearmod, group = tuition_type)) + 
  geom_line(aes(y = tuition_cost, color = tuition_type), size = 1) + 
  ggtitle("Historical Costs of College \n Adjusted for Inflation") + 
  theme(axis.text.x = element_text(angle = 45)) + 
  xlab("Year") + 
  ylab("College Tuition Cost") + 
  labs(color = "Type of Degree")+
  stat_smooth(method = lm, aes(x=yearmod,y=tuition_cost))
  
ggplotly(ts_plot)
```

```{r}
## replace _ with ,
tuition_income$income_lvl <- replace(tuition_income$income_lvl,tuition_income$income_lvl=="48_001 to 75,000","48,001 to 75,000")
```

```{r}
  #boxplot of income levels and net costs 
p <- ggplot(tuition_income, aes(x = income_lvl, y = net_cost, fill = income_lvl)) +
  geom_boxplot()+
  labs(x = "Income Levels", y="Net Cost", title = "Income Levels and Net Costs Compared", fill = "Income Level")+
  theme(axis.text.x = element_text(angle = 45)) 
ggplotly(p)
```

```{r}
## Merge salary potential and tuition cost datasets by name of college
potential_salary <- merge(salary_potential,tuition_cost,by=c("name"))
```

```{r}
  #ggplot for tuition vs career pay 
p <- ggplot(potential_salary, aes(x = in_state_tuition, y = mid_career_pay, fill = type,label = name)) +
  geom_point()+
  labs(x = "In State Tuition Cost", y="Estimated Mid Career Pay", title = "In State vs. Estimated Mid Career Pay for Types of Instituitions Costs Compared", fill = "Type of College") 
ggplotly(p)
```

```{r}
## Group in state tution into differnt levels
potential_salary <- potential_salary %>% 
  mutate(salary_group = case_when(
    between(in_state_tuition, 0, 10000) ~ "0 to 10000",
    between(in_state_tuition, 10000, 20000) ~ "10000 to 20000",
    between(in_state_tuition, 20000, 30000) ~ "20000 to 30000",
    between(in_state_tuition, 30000, 40000) ~ "30000 to 40000",
    between(in_state_tuition, 40000, 50000) ~ "40000 to 50000",
    between(in_state_tuition, 50000, 60000) ~ "50000 to 60000",
    between(in_state_tuition, 60000, 70000) ~ "60000 to 70000",
    TRUE ~ NA_character_
  ))
```

```{r}
p <- ggplot(potential_salary, aes(x = salary_group, y = mid_career_pay, fill = type)) +
  geom_boxplot()+
  labs(x = "In State Tuition", y="Estimated Mid Career Pay", title = "In State Tuition and Mid Career Pay Compared", fill = "Type of College")+
  theme(axis.text.x = element_text(angle = 45))
ggplotly(p)
```

```{r}
cor(potential_salary$in_state_tuition,potential_salary$mid_career_pay)
cor(potential_salary$in_state_tuition,potential_salary$early_career_pay)
```

```{r}
## Group in STEM percent into differnt levels
potential_salary <- potential_salary %>% 
  mutate(stem_group = case_when(
    between(stem_percent, 0, 10) ~ "0 to 10",
    between(stem_percent, 10, 20) ~ "10 to 20",
    between(stem_percent, 20, 30) ~ "20 to 30",
    between(stem_percent, 30, 40) ~ "30 to 40",
    between(stem_percent, 40, 50) ~ "40 to 50",
    between(stem_percent, 50, 60) ~ "50 to 60",
    between(stem_percent, 60, 70) ~ "60 to 70",
    between(stem_percent, 70, 80) ~ "70 to 80",
    between(stem_percent, 80, 90) ~ "80 to 90",
    between(stem_percent, 90, 100) ~ "90 to 100",
    TRUE ~ NA_character_
  ))
```

```{r}
p <- ggplot(potential_salary, aes(x = stem_group, y = mid_career_pay, fill = type)) +
  geom_boxplot()+
  labs(x = "STEM Percent", y="Estimated Mid Career Pay", title = "STEM Percent and Mid Career Pay Compared", fill = "Type of School")+
  theme(axis.text.x = element_text(angle = 45))
ggplotly(p)
```

```{r}
## Group in STEM percent into differnt levels
potential_salary <- potential_salary %>% 
  mutate(makebetter_group = case_when(
    between(make_world_better_percent, 0, 10) ~ "0 to 10",
    between(make_world_better_percent, 10, 20) ~ "10 to 20",
    between(make_world_better_percent, 20, 30) ~ "20 to 30",
    between(make_world_better_percent, 30, 40) ~ "30 to 40",
    between(make_world_better_percent, 40, 50) ~ "40 to 50",
    between(make_world_better_percent, 50, 60) ~ "50 to 60",
    between(make_world_better_percent, 60, 70) ~ "60 to 70",
    between(make_world_better_percent, 70, 80) ~ "70 to 80",
    between(make_world_better_percent, 80, 90) ~ "80 to 90",
    between(make_world_better_percent, 90, 100) ~ "90 to 100",
    TRUE ~ "Not Reported"
  ))
```

```{r}
p <- ggplot(potential_salary, aes(x = makebetter_group)) +
  geom_bar()+
  labs(x = "Feel Make World Better Percent", y="Frequency", title = "Histogram of Make World Better Percent")+
  theme(axis.text.x = element_text(angle = 45))
ggplotly(p)
```


```{r, echo = TRUE, message=FALSE, warning=FALSE}
  #read in more college data 
library(readr)
College_Data <- read_csv("College_Data.csv")
```

```{r, echo = TRUE, message=FALSE, warning=FALSE}

  #data wrangling
  
  #select variables 
college_df <- College_Data %>% 
  select(INSTNM, ADM_RATE, ACTCMMID, SAT_AVG, INEXPFTE, C150_4,ICLEVEL, MD_EARN_WNE_P6, DEBT_MDN, GRAD_DEBT_MDN)

college_df$ADM_RATE <- as.double(college_df$ADM_RATE)
college_df$ACTCMMID <- as.double(college_df$ACTCMMID)
college_df$SAT_AVG <- as.double(college_df$SAT_AVG)
college_df$INEXPFTE <- as.double(college_df$INEXPFTE)
college_df$C150_4 <- as.double(college_df$C150_4)
college_df$MD_EARN_WNE_P6 <- as.double(college_df$MD_EARN_WNE_P6)
college_df$DEBT_MDN <- as.double(college_df$DEBT_MDN)
college_df$GRAD_DEBT_MDN <- as.double(college_df$GRAD_DEBT_MDN)

  #update name of college for joining 
colnames(college_df)[colnames(college_df) == "INSTNM"] <- "name"

  #select variables 
col1 <- tuition_cost %>%
  select(c(name, room_and_board, in_state_tuition, out_of_state_tuition))
col2<- salary_potential %>%
  select(c(mid_career_pay, make_world_better_percent, stem_percent, name))

  #joing datasets 
col <- inner_join(col1, col2, by = "name")
college_factors <- inner_join(col, college_df, by = "name")
```


```{r}
  #view summary of data 
summary(college_factors)
```

```{r}
  #take out unnesseary values 
college_pc <- college_factors %>%
  select(-c(name, ICLEVEL))

  #remove NA and replace with 0s 
college_pc[is.na(college_pc)] <- 0
principle_components <- prcomp(college_pc, scale. = T)

  #look at principle compenents
principle_components$rotation[1:5,1:4]
std_dev <- principle_components$sdev
pr_var <- std_dev^2
prop_varex <- pr_var/sum(pr_var)

  #plot changing variance 
plot(cumsum(prop_varex), xlab = "Principle Component", ylab = "Cummulative Proportion of Variance Explained", type = "b")
abline(h = .95, col = "grey")
sum(prop_varex[1:9])
```
With 9 principle components, 95% of the variance is explained. 

```{r}
  #get principle components
college_pc <- cbind(college_pc, principle_components$x)
college_pc <- as.data.frame(college_pc[, 15:23])
head(college_pc)
```

```{r}

set.seed(2345)
  #create formula 
kmean_withinss <- function(k) {
    cluster <- kmeans(college_pc, k)
    return (cluster$tot.withinss)
}
  
  #look at best number of clusters 
max_k <- 20 
wss <- sapply(2:max_k, kmean_withinss)
elbow <- data.frame(2:max_k, wss)

ggplot(elbow, aes(x = X2.max_k, y = wss)) +
    geom_point() +
    geom_line() +
    scale_x_continuous(breaks = seq(1, 20, by = 1))
```

We will choose 8 as the optimal cluster. 

```{r}
  #examine clusters 
pc_cluster <- kmeans(college_pc, 8)
pc_cluster$cluster
pc_cluster$centers
pc_cluster$size 
```

```{r}

  #correlation matix
center <- pc_cluster$centers
cluster <- c(1:8)
centerdf <- data.frame(cluster, center)
head(centerdf)

center_reshape <- gather(centerdf, features, values, PC1:PC9)

ggplot(data = center_reshape, aes(x = features, y = cluster, fill = values)) +
    scale_y_continuous(breaks = seq(1, 7, by = 1)) +
    geom_tile() +
    coord_equal() +
    theme_classic()
```

